---
title: "Introduction_1_loadmetadata"
author: "Hangjia Zhao"
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
  toc: true
vignette: >
  %\VignetteIndexEntry{Introduction_1_loadmetadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

[Progenetix](https://progenetix.org/) is an open data resource that provides curated individual cancer copy number variation (CNV) profiles along with associated metadata sourced from published oncogenomic studies and various data repositories. This vignette provides a comprehensive guide on accessing and utilizing metadata for samples or their corresponding individuals within the Progenetix database. 

If your focus lies in cancer cell lines, you can access data from [*cancercelllines.org*](https://cancercelllines.org/) by specifying the `dataset` parameter as "cancercelllines" in `pgxLoader` function. This data repository originates from CNV profiling data of cell lines initially collected as part of Progenetix and currently includes additional types of genomic mutations.

# Load library 

```{r setup}
library(pgxRpi)
```

## `pgxLoader` function 

This function loads various data from `Progenetix` database via the Beacon v2 API with some extensions (BeaconPlus).

The parameters of this function used in this tutorial:

* `type`: A string specifying output data type. Available options are "biosamples", "individuals", "analyses".   
* `filters`: Identifiers used in public repositories, bio-ontology terms, or custom terms such as c("NCIT:C7376", "PMID<!-- -->:22824167"). When multiple filters are used, they are combined using AND logic.
* `biosample_id`: Identifiers used in the query database for identifying biosamples. 
* `individual_id`: Identifiers used in the query database for identifying individuals. 
* `codematches`: A logical value determining whether to exclude samples from child concepts 
of specified filters in the ontology tree. If TRUE, only samples exactly matching the specified filters will be included. 
Do not use this parameter when `filters` include ontology-irrelevant filters such as PMID and cohort identifiers.
Default is FALSE.
* `limit`: Integer to specify the number of returned profiles. Default is 0 (return all). 
* `skip`: Integer to specify the number of skipped profiles. E.g. if skip = 2, limit=500, the first 2*500 =1000 profiles are skipped and the next 500 profiles are returned. Default is NULL (no skip). 
* `domain`: A string specifying the domain of the query data resource. Default is "http://progenetix.org".
* `dataset`: A string specifying the dataset to query. When the parameter `domain` is "http://progenetix.org", available options are "progenetix" (by defualt) and "cancercelllines".
* `entry_point`: A string specifying the entry point of the Beacon v2 API. Default is "beacon", resulting in the endpoint being "http://progenetix.org/beacon".

# Retrieve biosamples information

## Search by filters

Filters are a significant enhancement to the [Beacon](https://www.ga4gh.org/product/beacon-api/) query API, providing a mechanism for specifying rules to select records based on their field values. To learn more about how to utilize filters in Progenetix, please refer to the [documentation](https://docs.progenetix.org/common/beacon-api/#filters-filters-filtering-terms).

The `pgxFilter` function helps access available filters used in Progenetix by default. It is also possible to query available filters used in other resources via the Beacon v2 API by setting the `domain` and `entry_point` parameters accordingly. 
Here is an example usage:

```{r}
# access all filters
all_filters <- pgxFilter()
head(all_filters)
```

```{r}
# get all prefix
all_prefix <- pgxFilter(return_all_prefix = TRUE)
all_prefix
```

```{r}
# access specific filters based on prefix
ncit_filters <- pgxFilter(prefix="NCIT")
head(ncit_filters)
```

The following query retrieves metadata in Progenetix related to all samples of retinoblastoma, utilizing a specific filter based on an [NCIt code](https://ncit.nci.nih.gov) as a disease identifier.

```{r}
biosamples <- pgxLoader(type="biosamples", filters = "NCIT:C7541")
# data looks like this
biosamples[1:5,]
```

The data contains many columns representing different aspects of sample information.

## Search by biosample id and individual id 

In Progenetix, biosample id and individual id serve as unique identifiers for biosamples and the corresponding individuals. You can obtain these IDs through metadata search with filters as described above, or through [website](https://progenetix.org/search) interface query.

```{r}
biosamples_2 <- pgxLoader(type="biosamples", biosample_id = "pgxbs-kftvgioe",individual_id = "pgxind-kftx28q5")

biosamples_2
```

It's also possible to query by a combination of filters, biosample id, and individual id.

## Access a subset of samples

By default, it returns all related samples (limit=0). You can access a subset of them 
via the parameter `limit` and `skip`. For example, if you want to access the first 10 samples
, you can set `limit` = 10, `skip` = 0. 

```{r}
biosamples_3 <- pgxLoader(type="biosamples", filters = "NCIT:C7541",skip=0, limit = 10)
# Dimension: Number of samples * features
print(dim(biosamples))
print(dim(biosamples_3))
```

## Parameter `codematches` use

Some filters, such as NCIt codes, are hierarchical. As a result, retrieved samples may 
include not only the specified filters but also their child terms.

```{r}
unique(biosamples$histological_diagnosis_id)
```

Setting `codematches` as TRUE allows this function to only return biosamples that exactly match the specified filter, excluding child terms.

```{r}
biosamples_4 <- pgxLoader(type="biosamples", filters = "NCIT:C7541",codematches = TRUE)
unique(biosamples_4$histological_diagnosis_id)
```

To query the number of samples collected in Progenetix, you can use the `pgxCount` function. 

```{r}
pgxCount(filters = "NCIT:C7541")
```

# Retrieve individuals information

If you want to query metadata (e.g. survival data) of individuals where the samples 
of interest come from, set the parameter `type` to "individuals" and follow the same steps as above.

```{r}
individuals <- pgxLoader(type="individuals",individual_id = "pgxind-kftx26ml",filters="NCIT:C7541")
# data looks like this
individuals[173:174,]
```

# Retrieve analyses information 

If you want to know more details about data analyses, set the parameter `type` to "analyses". 
The other steps are the same, except the parameter `codematches` is not available because 
analyses data do not include filter information, even though it can be searched by filters.

```{r}
analyses <- pgxLoader(type="analyses",biosample_id = c("pgxbs-kftvik5i","pgxbs-kftvik96"))

analyses
```

# Visualization of survival data

Suppose you want to investigate whether there are survival differences associated with a particular disease, for example, between younger and older patients, or based on other variables. You can query and visualize the relevant information using the `pgxMetaplot` function.

## `pgxMetaplot` function

This function generates a survival plot using metadata of individuals obtained by the `pgxLoader` function.

The parameters of this function:

* `data`: The object returned by the `pgxLoader` function, which includes survival data about individuals.
* `group_id`: A string specifying which column is used for grouping in the Kaplan-Meier plot.
* `condition`: A string for splitting individuals into younger and older groups, following the ISO 8601 duration format. 
Only used if `group_id` is "age_iso".
* `return_data`: A logical value determining whether to return the metadata used for plotting. Default is FALSE.
* `...`: Other parameters relevant to KM plot. These include `pval`, `pval.coord`, `pval.method`, `conf.int`, `linetype`, and `palette` (see ggsurvplot function from survminer package)

### Example usage 

```{r}
# query metadata of individuals with lung adenocarcinoma
luad_inds <- pgxLoader(type="individual",filters="NCIT:C3512")
# use 65 years old as the splitting condition
pgxMetaplot(data=luad_inds, group_id="age_iso", condition="P65Y", pval=TRUE)
```

It's noted that not all individuals have available survival data. If you set `return_data` to TRUE, 
the function will return the metadata of individuals used for the plot.

# Session Info

```{r echo = FALSE}
sessionInfo()
```


