---
title: "Introduction_1_loadmetadata"
author: "Hangjia Zhao"
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
  toc: true
vignette: >
  %\VignetteIndexEntry{Introduction_1_loadmetadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

[Progenetix](https://progenetix.org/) is an open data resource that provides curated individual cancer copy number aberrations (CNA) profiles along with associated metadata sourced from published oncogenomic studies and various data repositories. This vignette provides a comprehensive guide on accessing and utilizing metadata for samples or their corresponding individuals within the Progenetix database.

# Load library 

```{r setup}
library(pgxRpi)
```

## `pgxLoader` function 

This function loads various data from `Progenetix` database.  

The parameters of this function used in this tutorial:

* `type` A string specifying output data type. Available options are "biosample", "individual", "variant" or "frequency".   
* `filters` Identifiers for cancer type, literature, cohorts, and age such as 
c("NCIT:C7376", "pgx:icdom-98353", "PMID<!-- -->:22824167", "pgx:cohort-TCGAcancers", "age:>=P50Y"). For more information about filters, see the [documentation](https://docs.progenetix.org/beaconplus/#filters-filters-filtering-terms).
* `filterLogic` A string specifying logic for combining multiple filters when query metadata. Available options are "AND" and "OR". Default is "AND". An exception is filters associated with age that always use AND logic when combined with any other filter, even if filterLogic = "OR", which affects other filters. 
* `individual_id` Identifiers used in Progenetix database for identifying individuals. 
* `biosample_id` Identifiers used in Progenetix database for identifying biosamples. 
* `codematches` A logical value determining whether to exclude samples 
from child concepts of specified filters that belong to cancer type/tissue encoding system (NCIt, icdom/t, Uberon). 
If TRUE, retrieved samples only keep samples exactly encoded by specified filters. 
Do not use this parameter when `filters` include cancer-irrelevant filters such as PMID and cohort identifiers.
Default is FALSE.
* `limit` Integer to specify the number of returned samples/individuals/coverage profiles for each filter. 
Default is 0 (return all). 
* `skip` Integer to specify the number of skipped samples/individuals/coverage profiles for each filter. 
E.g. if skip = 2, limit=500, the first 2*500 =1000 profiles are skipped and the next 500 profiles are returned. 
Default is NULL (no skip). 

# Retrieve meatdata of samples 

## Relevant parameters 

type, filters, filterLogic, individual_id, biosample_id, codematches, limit, skip

## Search by filters

Filters are a significant enhancement to the [Beacon](https://www.ga4gh.org/product/beacon-api/) query API, providing a mechanism for specifying rules to select records based on their field values. To learn more about how to utilize filters in Progenetix, please refer to the [documentation](https://docs.progenetix.org/beaconplus/#filters-filters-filtering-terms).

The `pgxFilter` function helps access available filters used in Progenetix. Here is the example use:

```{r}
# access all filters
all_filters <- pgxFilter()
# get all prefix
all_prefix <- pgxFilter(return_all_prefix = TRUE)
# access specific filters based on prefix
ncit_filters <- pgxFilter(prefix="NCIT")
head(ncit_filters)
```

The following query is designed to retrieve metadata in Progenetix related to all samples of lung adenocarcinoma, utilizing a specific type of filter based on an [NCIt code](https://ncit.nci.nih.gov) as an ontology identifier.

```{r}
biosamples <- pgxLoader(type="biosample", filters = "NCIT:C3512")
# data looks like this
biosamples[c(1700:1705),]
```

The data contains many columns representing different aspects of sample information. 

## Search by biosample id and individual id 

In Progenetix, biosample id and individual id serve as unique identifiers for biosamples and the corresponding individuals. You can obtain these IDs through metadata search with filters as described above, or through [website](https://progenetix.org/search) interface query.


```{r}
biosamples_2 <- pgxLoader(type="biosample", filters = "NCIT:C3512",
                        biosample_id = c("pgxbs-kftva5zv","pgxbs-kftva5zw"),
                        individual_id = "pgxind-kftx26ml")

biosamples_2[biosamples_2$biosample_id == "pgxbs-kftva5zv",c(1:7)]
```

## Access a subset of samples

By default, it returns all related samples (limit=0). You can access a subset of them 
via the parameter `limit` and `skip`. For example, if you want to access the first 1000 samples
, you can set `limit` = 1000, `skip` = 0. 

```{r}
biosamples_3 <- pgxLoader(type="biosample", filters = "NCIT:C3512",skip=0, limit = 1000)
# Dimension: Number of samples * features
print(dim(biosamples))
print(dim(biosamples_3))
```

## Query the number of samples in Progenetix

The number of samples in specific group can be queried by  `pgxCount` function.

```{r}
pgxCount(filters = "NCIT:C3512")
```

## Parameter `codematches` use

The NCIt code of retrieved samples doesn't only contain specified filters but contains child terms. 

```{r}
unique(biosamples$histological_diagnosis_id)
```

Setting `codematches` as TRUE allows this function to only return biosamples with exact match to the filter.

```{r}
biosamples_4 <- pgxLoader(type="biosample", filters = "NCIT:C3512",codematches = TRUE)

unique(biosamples_4$histological_diagnosis_id)
```

## Parameter `filterLogic` use

This function supports querying samples that belong to multiple filters. For example, If you want to retrieve information about lung adenocarcinoma samples from the literature 
PMID:24174329, you can specify multiple matching filters and set `filterLogic` to "AND".

```{r}
biosamples_5 <- pgxLoader(type="biosample", filters = c("NCIT:C3512","PMID:24174329"), 
                          filterLogic = "AND")
```

# Retrieve meatdata of individuals 

If you want to query metadata (e.g. survival data) of individuals where the samples 
of interest come from, you can follow the tutorial below.

## Relevant parameters  

type, filters, filterLogic, individual_id, biosample_id, codematches, limit, skip 

## Search by filters

```{r}
individuals <- pgxLoader(type="individual",filters="NCIT:C3270")
# Dimension: Number of individuals * features
print(dim(individuals))
# data looks like this
individuals[c(36:40),]
```

## Search by biosample id and individual id 

You can get the id from the query of samples

```{r}
individual <- pgxLoader(type="individual",individual_id = "pgxind-kftx26ml", biosample_id="pgxbs-kftvh94d")

individual
```

or by all available identifiers 

```{r}
total.individuals <- pgxLoader(type="individual",filters="NCIT:C3270",
                               individual_id = "pgxind-kftx26ml",
                               biosample_id= "pgxbs-kftvh94d")

dim(total.individuals)
```

# Session Info

```{r echo = FALSE}
sessionInfo()
```


