---
title: "Introduction_3_loadfrequency"
author: "Hangjia Zhao"
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
  toc: true
vignette: >
  %\VignetteIndexEntry{Introduction_3_loadfrequency}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

[Progenetix](https://progenetix.org/) is an open data resource that provides curated individual cancer copy number aberrations (CNA) profiles along with associated metadata sourced from published oncogenomic studies and various data repositories. This vignette offers a comprehensive guide on accessing and visualizing CNV frequency data within the Progenetix database. CNV frequency is pre-calculated based on CNV segment data in Progenetix and reflects the CNV pattern in a cohort. It is defined as the percentage of samples showing a CNV for a genomic region (1MB-sized genomic bins in this case) over the total number of samples in a cohort specified by filters.


# Load library 

```{r setup}
library(pgxRpi)
library(SummarizedExperiment)
library(GenomicRanges)
```

## `pgxLoader` function 

This function loads various data from `Progenetix` database.  

The parameters of this function used in this tutorial:

* `type` A string specifying output data type. Available options are "biosample", "individual",
"variant" or "frequency". 
* `output` A string specifying output file format. When the parameter `type` is "frequency", available options are "pgxfreq" or "pgxmatrix" .
* `filters` Identifiers for cancer type, literature, cohorts, and age such as 
c("NCIT:C7376", "pgx:icdom-98353", "PMID<!-- -->:22824167", "pgx:cohort-TCGAcancers", "age:>=P50Y").
For more information about filters, see the [documentation](https://docs.progenetix.org/beaconplus/#filters-filters-filtering-terms).
* `codematches` A logical value determining whether to exclude samples 
from child concepts of specified filters that belong to cancer type/tissue encoding system (NCIt, icdom/t, Uberon). 
If TRUE, retrieved samples only keep samples exactly encoded by specified filters. 
Do not use this parameter when `filters` include cancer-irrelevant filters such as PMID and cohort identifiers.
Default is FALSE.

# Retrieve CNV frequency data

## Relevant parameters

type, output, filters, codematches

## The first output format (`output` = "pgxfreq")

```{r}
freq_pgxfreq <- pgxLoader(type="frequency", output ='pgxfreq',
                         filters=c("NCIT:C4038","pgx:icdom-85003"))

freq_pgxfreq
```

The returned data is stored in [*GRangesList*](https://bioconductor.org/packages/release/bioc/html/GenomicRanges.html) container which consists of multiple *GRanges* objects.
Each *GRanges* object stores CNV frequency from samples pecified by a particular filter. 
Within each GRanges object, you can find annotation columns "gain_frequency" and "loss_frequency" in each row, which express the percentage values across samples (%) for gains and losses that overlap the corresponding genomic interval.

These genomic intervals are derived from the partitioning of the entire genome (GRCh38). Most of these bins have a size of 1MB, except for a few bins located near the telomeres. In total, there are 3106 intervals encompassing the genome.

To access the CNV frequency data from specific filters, you could access like this

```{r}
freq_pgxfreq[["NCIT:C4038"]]
```

To get metadata such as count of samples used to calculate frequency, use `mcols`:

```{r}
mcols(freq_pgxfreq)
```

The parameter `codematches` determines whether the calculation of CNV frequency excludes samples from child terms.

## The second output format (`output` = "pgxmatrix")

Choose 8 NCIT codes of interests that correspond to different tumor types

```{r}
code <-c("C3059","C3716","C4917","C3512","C3493","C3771","C4017","C4001")
# add prefix for query
code <- sub(".",'NCIT:C',code)
```

load data with the specified code

```{r}
freq_pgxmatrix <- pgxLoader(type='frequency',output = 'pgxmatrix',filters=code)
freq_pgxmatrix
```

The returned data is stored in [*RangedSummarizedExperiment*](https://bioconductor.org/packages/release/bioc/html/SummarizedExperiment.html) object, which is a matrix-like container where rows represent ranges of interest (as a GRanges object) and columns represent filters. 

To access the CNV frequency matrix, use `assay` accesssor

```{r}
head(assay(freq_pgxmatrix))
```

The matrix has 6212 rows (genomic regions) and 8 columns (filters). The rows comprised 3106 intervals with “gain status” plus 3106 intervals with “loss status”. 

The value is the percentage of samples from the corresponding filter having one or more CNV events in the 
specific genomic intervals. You could get the interval information by `rowRanges`

```{r}
rowRanges(freq_pgxmatrix)
```

For example, if the value in the second row and first column is 8.457, it means that 8.457% samples from the corresponding filter NCIT:C3059 having one or more duplication events in the genomic interval in chromosome 1: 400000-1400000.

Note: it is different from CNV status matrix introduced in Introduction_2_loadvariants. Value in this matrix is **percentage (%) of samples** having one or more CNVs overlapped with the binned interval while the value in CNV status matrix is **fraction in individual samples** to indicate how much the binned interval overlaps with one or more CNVs in the individual sample.

# Visualization of CNV frequency data using 'pgxfreq' format

## `pgxFreqplot` function

This function provides CNV frequency plots by genome or chromosomes as you request.

The parameters of this function:

* `data`: frequency object returned by `pgxLoader` function. 
* `chrom`: a vector with chromosomes to be plotted. If NULL, return the plot by genome. If specified the frequencies are plotted with one panel for each chromosome. Default is NULL.
* `layout`: number of columns and rows in plot. Only used in plot by chromosome. Default is c(1,1). 
* `filters`: Index or string value to indicate which filter to be plotted. The length of filters is limited to one if the parameter `circos` is False. Default is the first filter.
* `circos`: a logical value to indicate if return a circos plot. If TRUE, it can 
return a circos plot with multiple group ids for display and comparison. Default is FALSE.
* `highlight`: Indices of genomic bins to be highlighted with red color.
* `assembly`: A string specifying which genome assembly version should be applied to CNV frequency plotting. Allowed options are "hg19", "hg38". Default is "hg38" (genome version used in Progenetix).

## CNV frequency plot by genome 

### Input is `pgxfreq` object.

```{r, fig.width=7, fig.height=5}
pgxFreqplot(freq_pgxfreq, filters="pgx:icdom-85003")
```

### Input is `pgxmatrix` object. 

```{r, fig.width=7, fig.height=5}
pgxFreqplot(freq_pgxmatrix, filters = "NCIT:C3512")
```

## CNV frequency plot by chromosomes

```{r, fig.width=7, fig.height=5}
pgxFreqplot(freq_pgxfreq, filters='NCIT:C4038',chrom=c(1,2,3), layout = c(3,1))  
```

## CNV frequency circos plot 

```{r,fig.width=6, fig.height=6}
pgxFreqplot(freq_pgxfreq, filters='pgx:icdom-85003', circos = TRUE)
```

The circos plot also supports multiple group comparison

```{r,fig.width=6, fig.height=6}
pgxFreqplot(freq_pgxfreq,filters= c("NCIT:C4038","pgx:icdom-85003"),circos = TRUE) 
```

## Highlight interesting genomic intervals

If you want to look at the CNV frequency at specific genomic bins, you can use `highlight` parameter.
For example, when you are interested in CNV pattern of CCND1 gene in samples with infiltrating duct carcinoma (icdom-85003).
You could first find the genomic bin where CCND1 (chr11:69641156-69654474) is located.

```{r}
# Extract the CNV frequency data frame of samples from 'icdom-85003' from 
# the previously returned object
freq_IDC <- freq_pgxfreq[['pgx:icdom-85003']]
# search the genomic bin where CCND1 is located
bin <- which(seqnames(freq_IDC) == 11 & start(freq_IDC) <= 69641156 &  
             end(freq_IDC) >= 69654474)
freq_IDC[bin,]
```

Then you could highlight this genomic bin like this

```{r, fig.width=7, fig.height=5}
pgxFreqplot(freq_pgxfreq,filters = 'pgx:icdom-85003', chrom = 11,highlight = bin)
```

Note: For CNV analysis of specific genes, the highlighted plot is rough as a reference, 
because the bin size in frequency plots is 1MB, which is possible to cover multiple genes.

The highlighting is also available for genome plots and circos plots. And you could highlight 
multiple bins by a vector of indices.

```{r, fig.width=7, fig.height=5}
pgxFreqplot(freq_pgxfreq,filters = 'pgx:icdom-85003',highlight = c(1:100))
```

# Session Info

```{r echo = FALSE}
sessionInfo()
```

